---
title: "Final Exam"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

Please read all of the questions and tables carefully and follow all instructions. Each question has an allotted point value. Be as thorough and specific as possible; extra calculations and incorrect information, even in the presence of correct information, will result in point deductions. Be sure to show all formulas and all necessary work to answer the questions. You may upload your completed exam to the Elms course site.

::: callout-note
While this is an open-note exam, you are not to receive assistance from anyone else. As usual, the Honor Code applies.
:::

**Total points: 50**

## Conceptual Questions

Please include all work (and computations) necessary to answer the questions.

**Total points: 20 (2 points each)**

### Question 1

Suppose you estimate a bivariate regression model with a total sample size of 50 and obtain a standard error for $\beta_1$ of 2.50. What is the smallest regression coefficient ($\beta_1$) that you could have and still be able to reject the null hypothesis (of no relationship between $X_1$ and $Y$) at the 0.05 level with a one-tailed test?

**Answer:**

In a one-tail T test at the .05 level, with 48 degrees of freedom (50 -2) and in a bivariate regression model the critical t value is approximately 1.68. To reject the null hypothesis the absolute value of T statistic must be greater than the critical value. Using the formula t = B1/SE, we can solve for the value (1.68 \* 2.50 = 4.2), we get the smallest regression coefficient and still reject the null hypothesis at a .05 level would be 4.2 or -4.2 depending on the direction of the relationship.

### Question 2

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and find that $\beta_1 = 4.20$, $SE_{\beta_1} = 1.90$, $t = 2.21$, $p = 0.01$, and a 95% confidence interval is \[0.48, 7.92\]. What is my best estimate of the effect of a two-unit change in $X$ on the mean of $Y$ in the population?

**Answer:**

Given that B1 is is 4.20 this means that for one unit increase in X1 we can expect on average a 4.20 unit increase in Y. Since we are looking at a two unit change we can estimate that on average there is a 8.40 (4.20\*2) unit change in X1 on the means of Y.

### Question 3

I conduct an OLS regression with a sample size of 90 and 5 independent variables. To determine a p-value for each coefficient, I would examine a t-distribution with how many degrees of freedom?

**Answer:**

90 - 6 (5 degrees of freedom + 1 intercept) = 84 degrees of freedom.

### Question 4

Suppose I regress $Y$ on $X$ and compute the mean response for $Y$ at some specified value of $X$. When determining the confidence interval around this mean response, which of the following will **NOT** have any effect on the width (or, size) of that confidence interval?

A.  Total sample size,

B.  Mean-squared error,

C.  The specified value of $X$,

D.  **None of the above: all of these (above) will affect the confidence interval,**

E.  There is not enough information to answer this question.

The total sample size can affect the precision of the estimate as larger sample size lead to narrower confidence interval. The mean squared error is related to the variability of the data and the specific value of x can affect the precision of the estimate. So all those factors will affect the size of the confidence interval.

### Question 5

I regress $Y$ on $X$ and find that $\beta_1$ has a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  The lower and upper bound of a 95% confidence interval around $\beta_1$ will have the same sign,

B.  The absolute value for the t-statistic for $\beta_1$ will be greater than 1.96,

C.  A 90% confidence interval around $\beta_1$ will not contain zero,

D.  **All of the above,**

E.  None of the above and/or there is not enough information.

If the confidence interval does not include zero (since the p-values is less than .05) then the upper and lower bounds will have the same sign. In a two-tailed test, a p-value of 0.04 corresponds to a t-statistic whose absolute value is greater than the critical value for the 5% significance level (which is approximately 1.96). A narrower confidence interval (like the 90% confidence interval) will also not contain zero. Therefore the answer is all of above.

### Question 6

I regress $Y$ on three independent variables -- $X_1$, $X_2$, and $X_3$ -- and I find the following 95% confidence intervals -- $\beta_1$: \[0.12, 1.45\], $\beta_2$: \[-0.01, 0.15\], and $\beta_3$: \[-0.64, -0.01\]. Which of the following statements is most accurate?

A.  $\beta_2$ and $\beta_3$ will have negative coefficients, and $\beta_1$ is statistically significant at the 0.05 level (two-tailed),

B.  $\beta_1$ will have a positive coefficient, and $\beta_1$ is the only statistically significant coefficient (of the three coefficients) at the 0.05 level (two-tailed),

C.  $\beta_1$ **will have a positive coefficient,** $\beta_2$ **and** $\beta_3$ **will have negative coefficients, and only** $\beta_1$ **and** $\beta_3$ **are statistically significant at the 0.05 level (two-tailed),**

D.  All of the above,

E.  None of the above and/or there is not enough information.

### Question 7

Suppose I estimate a regression with two independent variables and obtain a $R^2$ of 0.40 where the Residual Sum of Squares is equal to 5,150. What does the Total Sum of Squares equal in this regression model?

**Answer:**

T= RSS/ 1 - $R^2$, so 5150/1-0.4. Therefore the total sum of squares in this regression model is 8,583.

### Question 8

I regress $Y$ on $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$). I find that $\beta_1 = -0.87$ with a two-tailed p value of 0.001 and $\beta_2 = 3.52$ with a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  $X_2$ **has a larger substantive effect on** $Y$ **than** $X_1$**, but the effect of** $X_1$ **is more statistically significant,**

B.  $X_1$ has a larger substantive effect on $Y$ than $X_2$ and $X_1$ is more statistically significant than $X_2$,

C.  $X_1$ has a larger substantive effect on $Y$ than $X_2$, but the effect of $X_2$ is more statistically significant,

D.  None of the above and/or there is not enough information.

### Question 9

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and obtain a $R^2$ of 0.45. Then, I regress $Y$ on both $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$) and obtain a $R^2$ of 0.53. Which of the following statements must be true?

A.  The correlation between $X_1$ and $Y$ is stronger than the correlation between $X_2$ and $Y$,

B.  The coefficient on $\beta_1$ is statistically significant in both models,

C.  **The model with** $X_1$ **and** $X_2$ **explains more of the variation in** $Y$ **than a model with just** $X_1$**,**

D.  All of the above: all of these statements are true.

This is the correct statement evidenced by the increase in R2 from 0.45 to 0.53 when X2​ is included in the model, indicating that the model with both X1​ and X2​ explains more of the variance in Y than the model with just X1​.

### Question 10

Based on the following regression equation, might the negative coefficient on the South variable be the result of southern states having a lower proportion of high school graduates than non-southern states? In no more than one sentence, explain your answer.

$$
Turnout = 34 + 0.5(Percent\ High\ School\ Graduates) – 5.9(South)
$$

**Answer**: The negative coefficient on the South variable could be attribute to southern states having a lower proportion of high school graduates, if this factor is not accounted for elsewhere in the model.

## Applied Questions

All datasets referenced below are posted on the ELMs course site.

### Question 1

*Points: 15*

Use the `world` dataset to answer the following questions. The relevant variables and their coding information are as follows: `literacy` indicates a country's literacy rate; `dem_score14` represents a country's level of democratization (higher values indicate greater democratization); `spendeduc` reflects the amount of public expenditures on education as a percentage of GDP; `gdp_10_thou` represents GDP per capita (one unit is \$10,000 USD); `educ_quality` indicates the average quality rating of a country's educational system; and `ungr9095` represents the percent average annual population growth.

#### Part A

*Points: 4*

Evaluate the bivariate relationship (using R) between a country's support for public education and its literacy rate. Report the results in a professionally formatted table. Do countries that spend more money on education appear to have a higher literacy rate?

**Answer:**

```{r}
library(stargazer)
library(modelsummary)

world <- read.csv("world.csv")

model_world_1 <- lm(literacy ~ spendeduc, data = world)
summary(model_world_1)


world_titles <- c("Education Spending %", "Intercept")
stargazer(model_world_1,
          covariate.labels = world_titles,
          dep.var.labels = c("Literacy Rate"),
          ci = TRUE,
          ci.level = 0.95,
          type = "text")





```

Based on the regression results , there appears to be a positive relationship between education spending and literacy rates. The coefficient for Education Spending % is on average 1.722 and is statistically significant at the 10% level. This suggests that for each unit increase in Education Spending %, the Literacy Rate increases by 1.722 units, holding all other factors constant. Furthermore, the intercept of 73.270 suggests that when Education Spending % is zero, the expected Literacy Rate is 73.270.

#### Part B

*Points: 5*

Estimate a second regression model (using R) that includes both a country's public expenditures on education and its level of democratization as independent variables. When controlling for democratization, does education expenditures exhibit a significant impact on literacy? Report the results in a professionally formatted table. Be sure to discuss the extent to which the results change, and if so, why they changed.

```{r}


world <- read.csv("world.csv")


model_world_2 <- lm(literacy ~ spendeduc + dem_score14, data = world)
summary(model_world_2)

world_titles <- c("Education Spending %", "Democratization Score", "Intercept")
stargazer(model_world_2,
          covariate.labels = world_titles,
          dep.var.labels = c("Literacy Rate"),
          ci = TRUE,
          ci.level = 0.95,
          type = "text")




```

Based on this regression table educational spending does not exhibit a significant impact on literacy rates when controlling for the democratization scores. The coefficient of ed. spending is on average 0.258, but it's not statistically significant at the 10% level since its confidence interval (-1.314, 1.830) includes zero. This suggests that the percentage of education spending does not have a significant effect on the literacy rate, according to this model. Comparing this model with the previous one, we can see that the coefficient of education spending has increased from on average 0.258 to 1.722 in the first regression and has become more statistically significant. This could be due to the omission of the democratization score variable, which might have been suppressing the effect of educational spending on literacy rate in the previous model. In a phenomenon is known as the suppression effect which occurs in multiple regression analysis.

#### Part C

*Points: 6*

Now estimate a third model (using R) that includes all of the variables listed above. Report the regression results in a professionally formatted table and interpret each coefficient. When interpreting the impact of democratization using this regression model, present/utilize a visual representation (using R) of the expected change in literacy as a function of democratization.

```{r}

library(ggplot2)
library(stargazer)

data <- read.csv("world.csv")

model <- lm(literacy ~ dem_score14 + spendeduc + gdp_10_thou + educ_quality + ungr9095, data = data)

labels <- c("Literacy Rate", "Democratization Score", "Educational Spending", "GDP", "Education Equality", "Population growth")

stargazer(model, title="Regression Results", align=TRUE, type = "text", covariate.labels = labels)


ggplot(data, aes(x = literacy, y = dem_score14)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Relationship between Democracy Score and Literacy",
    x = "Literacy",
    y = "Democracy Score (2014)"
  ) +
  theme_minimal()




```

-   A one-unit increase in the literacy is associated with an increase of approximately 2.546 units in literacy rates.

-   A one-unit increase in the Democratization Score is associated with an increase of approximately 0.112 units in literacy rates around the world while there is a positive relationship (as seen in the slope of the line) between the Democratization Score and literacy in this dataset, this relationship is not strong enough to confidently state that it exists in the population your sample represents.

-   A one-unit increase in Educational Spending is associated with an increase of approximately 2.360 units in literacy rates.

-   A one-unit increase in GDP is associated with an increase of approximately 3.812 units in literacy rates around the world.

-   A one-unit increase in Education Equality is associated with a decrease of approximately 3.326 units in literacy, and finaly, A one-unit increase in Population growth is associated with an increase of approximately 55.910 units in literacy

The results for each of the variables alone is under the condition that all other variables are held constant.

### Question 2

*Points: 15*

Using the `states` dataset, use R to regress the variable `obama08` (the percentage of a state's vote that President Obama received in the 2008 U.S. presidential election) on the following independent variables: `cig_tax` represents the amount of a state's cigarette tax (in dollars); `college` is the percentage of a state's population that graduated college (0 to 100); `union07` is the percentage of a state's workers that are union members (0 to 100); and `south` is a dichotomous variable coded as `1` if the state is located in the South (`0` otherwise). Report the results in a professionally formatted table. Answer the following questions.

```{r}

library(dplyr)
library(broom)

states <- read.csv("states.csv")

model <- lm(obama08 ~ cig_tax + college + union07 + south, data = states)

summary(model)

tidy_model <- broom::tidy(model)

names(tidy_model) <- c("Variable", "Estimate", "Standard Error", "Statistical Significance", "p-value")

tidy_model$Variable <- recode(tidy_model$Variable,
                              "cig_tax" = "Cigarette Tax",
                              "college" = "Percent Graduated College",
                              "union07" = "Union Membership",
                              "south" = "Southern State")

print(tidy_model)




```

#### Part A

*Points: 6*

Interpret the substantive results from this regression model. Be sure to discuss the direction, magnitude, and statistical significance of each slope coefficient. Does the `union07` variably exhibit a substantively significant impact on voting behavior in the 2008 election?

**Answer:**

Cigarette Tax For each unit increase in cig_tax, holding all other variables constant, obama08 is expected to increase by 0.5230. However, this is not statistically significant (p=0.815676), meaning we cannot confidently say that cig_tax has an effect on obama08.

College: For each unit increase in college, obama08 is expected to increase by 0.9646, holding all other variables constant. This is statistically significant at the 0.001 level (p=0.000299).

Union Membership: For each unit increase in union07, obama08 is expected to increase by 0.7514, holding all other variables constant. This is statistically significant at the 0.01 level (p=0.003511).

South: For each unit increase in south, obama08 is expected to increase by 1.3928, holding all other variables constant. However, this is not statistically significant (p=0.593179).

The variable union07 does exhibit a statistically significant impact on voting behavior in the 2008 election, as its p-value is less than 0.05. However, whether it is substantively significant would depend on the context and the specific research question being asked.

college and union07 are statistically significant predictors as their p-values are less than 0.05.

cig_tax and south are not statistically significant predictors as their p-values are greater than 0.05.

#### Part B

*Points: 2*

Interpret the intercept coefficient. What does this represent and is it meaningful to interpret this coefficient on its own? Why or why not?

**Answer:**

When all other variables are held at zero, the expected value of the obama08 variable is 16.1280. This is statistically significant at the 0.05 level (p=0.018374). Interpreting the intercept on its own can sometimes be misleading or not meaningful at all, especially if zero is not a meaningful value for the independent variables. Once more it does not make realistic sense that the cigarette tax would every possess a zero value in the real world, so interpreting it on its own wouldn't make much sense.

*Points: 2*

**Part C**

What is the expected mean percentage of Obama's vote total in a Southern state with a cigarette tax of \$2.00, a state population where 30% of people are college graduates, and where 15% of workers in a state are union members?

Expected Value = $Beta0$ + $Beta1$​ \* cig_tax + $Beta2$ ​× college + $Beta3$ × union07+ $Beta4$​ × south

Expected Value= 16.1280 + (0.5230 \* 2.00) + (0.9646 \* 0.30) + (0.7514 \* 0.15) + (1.3928\* 1)

Based on the math above, the expected mean toal percentage is 18.96 for Obama's election.

**Part D**

*Points: 3*

What is a 90% confidence interval around the coefficient for the `south` variable.

Lower Limit = 1.3928 - (1.684\*2.5885)

Upper Limit = 1.3928+(1.684\*2.5885)

So the 90% confidence interval around the south variable would be -2.9662 and 5.7518

#### Part E

*Points: 3*

What is the null hypotheses is that the F-test in this regression output is testing?

The null and alternative hypotheses for the regression are as follows:

-   Null Hypothesis (H0): All the regression coefficients are equal to zero, i.e., none of the predictors (cig_tax, college, union07, south) have an effect on the dependent variable.

-   Alternative Hypothesis (H1): At least one regression coefficient is not equal to zero, i.e., at least one of the predictors has a significant effect on the dependent variable.

In this case, the p-value associated with the F-statistic is 8.504e-07, which is less than 0.05, suggesting that we can reject the null hypothesis. This means that at least one of the predictors is statistically significant in the model.
