---
title: "Problem Set 6"
subtitle: "Due date: 30 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

```{R}

library(tidyverse)
library(dplyr)

```

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 40**

## Question 1

*Points: 10*

:::{.callout-tip}
9/10
:::

I hypothesize that the mean mathematics SAT score for University of Maryland students is different than 600. I take an SRS of 625 students and find that the mean mathematics score among those students is 610. The standard deviation of my sample is 75.

### Part A

:::{.callout-tip}
2/2
:::

Based on my sample, what is a point estimate for the mean mathematics SAT score for University of Maryland students?

The point estimate for the mean math Sat score for UMD is equal to the sample mean which is 610. So 610 is the point estimate.

### Part B

:::{.callout-tip}
1/2
:::

What is a 95% confidence interval around that estimate?

Sample Mean = 610

Standard Deviation = 75

Sample Size = 625

Z = 1.96

:::{.callout-tip}
We are working with a sample, so we have to use the t-statistic not the z-score. 
:::

Confidence Interval = 610 +/- (1.96 x 75)/√625

610 +/- 5.88

Lower Bound is 604.12 and Upper Bound is 615.88 for a 95% interval.

```{R}

sample_mean <- 610
standard_deviation <- 75
sample_size <- 625
confidence_level <- 0.95

SE <- standard_deviation / sqrt(sample_size)

z_score <- qnorm(1 - (1 - confidence_level) / 2)

ME <- z_score * SE

lower_bound <- sample_mean - ME
upper_bound <- sample_mean + ME

cat("95% Confidence Interval: [", round(lower_bound, 2), ", ", round(upper_bound, 2), "]\n")





```

### Part C

:::{.callout-tip}
2/2
:::

What is the null hypothesis?

The null hypothesis would be a mean mathematics SAT score 600.

### Part D

:::{.callout-tip}
2/2
:::

What is the two-tailed p-value and what does it mean?

Z = 610 -600/ 75√625 = 3.33

P Value = .0002

```{r}

sample_mean <- 610
sample_size <- 625
sample_sd <- 75
population_mean <- 600

se <- sample_sd/ sqrt(sample_size)

t_stat <- (sample_mean - population_mean)/ se

df <- sample_size -1 

p_value <- 2 * pt(-abs(t_stat), df)

cat("Two-Tailed P-Value:", p_value, "/n")



```

The two tailed P value is a measure that helps assess the probability of observing a test statistic as an extreme of what is observed and helps quantify the likelihood of the observed data under the null hypothesis.

### Part E

:::{.callout-tip}
2/2
:::

Does the data support my hypothesis? Why, or why not?

Since we got a p value that is less than the significance of the typical use .05 significance level we can reject the null hypothesis that the mean mathematics SAT score for University of Maryland students is equal to 600.

## Question 2

*Points: 10*

:::{.callout-tip}
9/10
:::

I take an SRS of 900 citizens in a public opinion survey and find that the mean feeling thermometer rating (on a 0 to 100 scale) of President Biden is 51.5 with a standard deviation of 25. Now, suppose I hypothesize that feelings toward Biden should be greater than 50.

### Part A

:::{.callout-tip}
1/2
:::

What is the null hypothesis?

The null hypothesis would be that the mean feeling thermometer rating for President Biden would be less than or equal to 50.

:::{.callout-tip}
The null hypothesis has to be a single point. Here it is that the mean feeling thermometer rating for President Biden is equal to 50.
:::

### Part B

:::{.callout-tip}
4/4
:::

What is a 95% confidence interval around the sample statistic? Does this interval indicate support for a two-tailed test of my hypothesis (`p < 0.05` as the threshold)? Why, or why not?

```{r}

sample_mean <- 51.5
sample_size <- 900
sample_sd <- 25

confidence_level <- 0.95

standard_error <- sample_sd / sqrt(sample_size)

margin_of_error <- qt((1 + confidence_level) / 2, df = sample_size - 1) * standard_error

confidence_interval <- c(sample_mean - margin_of_error, sample_mean + margin_of_error)

confidence_interval





```

Since the confidence interval does not include the value 50 , it suggests that there is some evidence that the mean feeling thermometer rating is greater than 50. In this case, the data supports the alternative hypothesis that feelings toward President Biden are greater than 50. This provides some indication that a two-tailed test of our hypothesis , with a significance level of 0.05, would likely lead to rejecting the null hypothesis in favor of the alternative hypothesis.

### Part C

:::{.callout-tip}
4/4
:::

What is a 90% confidence interval around the sample statistic? Does this interval indicate support for a one-tailed test of my hypothesis (`p < 0.05` as the threshold)? Why, or why not?

```{R}

sample_mean <- 51.5
sample_size <- 900
sample_sd <- 25

confidence_level <- 0.90  # Change this to 0.90 for a 90% confidence interval

standard_error <- sample_sd / sqrt(sample_size)

margin_of_error <- qt((1 + confidence_level) / 2, df = sample_size - 1) * standard_error

confidence_interval <- c(sample_mean - margin_of_error, sample_mean + margin_of_error)

confidence_interval






```

Since the confidence interval includes values on both sides of 50, so it does not support a two-tailed test (p \< 0.05) for the alternative hypothesis. It suggests that the data is consistent with the possibility that the true population mean is greater than 50, but it does not provide strong evidence that it is significantly greater.

## Question 3

*Points: 5*

:::{.callout-tip}
5/5
:::

Suppose I conduct a difference of means test and obtain a t-statistic of 2.50. What does this indicate about the statistical significance of the sample mean compared to the null hypothesis?

Answer:

The t - statistic of 2.50 shows the number of stand deviations the sample mean is away from the mean of the null hypothesis. The significance of this depends on the degrees of freedom and significance level chosen. If the value of t - statistic is greater than the critical t value than you would reject the null hypothesis and it would suggest that the difference in means in statistically significant.

## Question 4

*Points: 5*

:::{.callout-tip}
3.5/5
:::

I hypothesize that fewer than 40% of registered voters approve of President Biden. I conduct an SRS of 720 registered voters and find that 37.9% of them approve of President Biden.

### Part A

:::{.callout-tip}
1/1
:::

Based on my sample, what is a point estimate for the proportion of the population that approves of President Biden?

The point estimate is the sample proportion of 37.9% or 0.379

### Part B

:::{.callout-tip}
1/1
:::

What is a 90% confidence interval around that estimate?

.379 +/- 1.645(√0.379(1-0.379)/720))

SE = 0.0181

Lower Bound : .379 - 1.645 x 0.0181 = .349 or 34.9%

Upper Bound: .379 + 1.645 x 0.0181 = .408 or 40.8%

```{r}

n <- 720
p_hat <- 0.379

se <- sqrt((p_hat * (1 - p_hat)) / n)

z <- qnorm(0.95)

me <- z * se

ci_lower <- p_hat - me
ci_upper <- p_hat + me

cat("The 90% confidence interval is [", ci_lower, ", ", ci_upper, "]")






```

### Part C

:::{.callout-tip}
0/1
:::

What is the null hypothesis?

Null Hypothesis would be the proportion of registered voters who approve of President Biden is equal to or greater than 40%.

:::{.callout-tip}
Again, it needs to be a single point. 
:::

### Part D

:::{.callout-tip}
0.5/1
:::

What is the one-tailed p-value and what does it mean?

```{r}
n <- 720

p_observed <- 0.379  # 37.9%

p_null <- 0.4  # 40%

result <- prop.test(x = p_observed * n, n = n, p = p_null, alternative = "less")

p_value <- result$p.value

cat("One-tailed p-value:", p_value, "\n")

```

:::{.callout-tip}
Be careful when using functions that include hidden defaults. This function applies Yates' continuity correction by default, which inflates your p-value. The p-value we were looking for is: 

```{r}
prop.test(x = 0.379*720, n = 720, p = 0.4, alternative = "less", correct = F)$p.value
```

Always good to also do something by hand when using a new function to make sure that it is behaving as you expect it to. 
:::

The one tailed P value means that there is a 13.3% chance of observing a percentage of voters as 37.9%. Since we have a relativity high P value we likely don't have enough evidence to reject the null hypothesis.

### Part E

:::{.callout-tip}
1/1
:::

Does the data support my hypothesis? Why, or why not?

Given that the p-value suggests that the observed proportion of 37.9% is not significantly different from the hypothesized 40%. There is a relatively high probability of obtaining results by random chance alone, given the null hypothesis of being greater than 40%. Therefore, we should not conclude that the approval rate is significantly less than 40% based on this data and should not support the hypothesis.

## Question 5

*Points: 5*

:::{.callout-tip}
5/5
:::

I hypothesize that in countries where Islam is the predominate religious group, women have a greater number of children on average (i.e., the total fertility rate is higher). Use the `world` dataset to test my hypothesis. State the null hypothesis and interpret what the results tell us about the null and alternative hypotheses. Be sure to show all work necessary to find the answer (i.e., you may use R to assist you, but you should show the necessary computations by hand).

::: callout-note
The `world` data set can be found in `poliscidata::world`.
:::

```{r}
library(tidyverse)
library(dplyr)
# poliscidata::world

world <- poliscidata:: world |>
  mutate(group_religion = ifelse(religoin == "Muslim", "Muslim", "non_Muslim"))

world_religion_m <- world |>
  filter(group_religion%in%c("non_Muslim")) |>
  select(group_religion, fertility)

p_hat_5 <- 46/167
p_0_5 <- 119/167
sample_size_religion <- 167

z_t_stat_5 <- (p_hat_5 - p_0_5)/sqrt(p_0_5 * (1 - p_0_5)/ sample_size_religion)

p_value_5 <- pnorm(z_t_stat_5)

cat("One-tailed p-value:", p_value_5)
                                
                        


```

```         
One-tailed p-value: 4.675515e-36
```

The null hypothesis is that there is no significant difference in the total fertility rates where Islam is the dominant religion vs. countries where it is not the dominant religion. The alternative hypothesis being there is a significant difference in the total fertility rates between the countries that have Islam as their predominant religion and those that that don't.

When we calculate the one-tailed p value using we get the result of 4.675515e-36 and with such a small p value we would normally reject the null hypothesis. These results suggest that there is indeed a statistically significant difference in the total fertility rate between countries where Islam is the predominant religion and countries where it is not.
