---
title: "Problem Set 1"
subtitle: "Due date: 18 September"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 25**

::: {.callout-tip title="HG comments"}
Great work. Nice answers. 21/25.
:::

::: {.callout-tip title="HG comments"}
Make sure to render your .qmd document before you submit it. We need to be able to run this script with no errors. If you can render the document with no errors, you are sorted.

You have some stray ``` throughout this document. Remember, you only need to use those to open and close a code chunk. 
:::

::: {.callout-tip title="HG comments"}
You need to load all the packages you will use in your script to access their functions during an R session. You need to load these packages before you use them within a script. I recommend including a code chunk right at the start of your document that does this. Your document will not render unless you load all of your necessary packages. 

For example: 

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(poliscidata)
```
:::

## Question 1

*Points: 5*

::: {.callout-tip title="HG comments"}
Correct and good code. 5/5.
:::

Using the `gss` dataset (which has survey data with the individual respondent as the unit of analysis), create a frequency distribution and bar chart for each of the following variables: `degree` and `partyid_3`. Describe the distribution of each variable in detail.

::: callout-note
The `gss` dataset can be found in `poliscidata::gss`. Take a look at `?gss` to see what these variables measure.
:::

##### Answer 1 : Frequency Distribution & Bar Chart (degree)

```{r}
freqC(gss$degree) # Describing a Nominal-Level Variable
```
 
 The distribution of this variable is skewed to the right as the bulk of the observations (nearly half) occur at the high school level thereby shifting the center of of the distribution to the right. However in some regards the distribution appears bi modal as observations on the left increase with the number of bachelor and graduate degrees. 


##### Answer 1 : Frequency Distribution & Bar Chart (partyid_3)

```{r}

#Frequency Table (Party ID)

freq(gss$partyid_3,) # Describing a Nominal-Level Variable


```

This variable has a relatively symmetric distribution with a slight skew to the left. The center of the distribution is easy to locate and although it being located further to the left due to similarity in the number of democrats and independents.  

::: {.callout-tip title="HG comments"}
Be careful when talking about skew among categorical variables. This assumes an order that may not necessarily be apparent. Worth noting in your answer that you are thinking of these variables as Dem - Ind - Rep. 
:::

## Question 2

Examine the following vector of (fake) student IQ test scores:

```{r}
iq <- c(
  145, 139, 126, 122, 125, 130, 96, 110, 118, 118, 101, 142, 134, 124, 112, 109, 
  134, 113, 81, 113, 123, 94, 100, 136, 109, 131, 117, 110, 127, 124, 106, 124, 
  115, 133, 116, 102, 127, 117, 109, 137, 117, 90, 103, 114, 139, 101, 122, 105, 
  97, 89, 102, 108, 110, 128, 114, 112, 114, 102, 82, 101
)



```

### Part A

*Points: 5*

::: {.callout-tip title="HG comments"}
Correct and nicely described. 5/5.
:::

Find the five-number summary, mean, and standard deviation for these data. Also, are there any suspected outliers in the distribution? If so, what are they and how do you know?

```{r}

x <- c(145, 139, 126, 122, 125, 130, 96, 110, 118, 118, 101, 142, 134, 124, 112, 109, 134, 113, 81, 113, 123, 94, 100, 136, 109, 131, 117, 110, 127, 124, 106, 124, 115, 133, 116, 102, 127, 117, 109, 137, 117, 90, 103, 114, 139, 101, 122, 105,97, 89, 102, 108, 110, 128, 114, 112, 114, 102, 82, 101)

fivenum(x)
mean(x)
sd(x)
```
Five Number Summary
Min - 81.0
Q1 - 104.0
Median - 114.0
Q3 - 125.5
Max - 145.0
Mean - 114.9833
SD - 14.80093

We can determine the interquartile range and use calculate the upper and lower fence of the data. The interquatile range (Q3-Q1) is 21.5 and the upper and lower fence (Q3,1 +/- (1.5*IQR)) is 157.75 and 71.75 respectivly. As all the data falls within these two fences we can determine that there is not significant outliers in this particular dataset. 


### Part B

*Points: 5*

:::{.callout-tip title="HG comments"}
A good understanding of an important characteristic of normally distributed data, but does not answer the question asked. 2/5.
:::

In large populations, IQ scores are standardized to have a mean of 100 and a standard deviation of 15. In what way does the distribution among these students differ from the overall population?

Based on the rule of 68-95-99.7 we would expect 68% of the total population to fall between 85 and 115, 95% between 70 and 130, and 99.7% less than 55 or higher than 145. Based on these prediction we can see that the group of students generally falls within 68% of the total population and follows the  distribution of the general population.

::: {.callout-tip title="HG comments"}
A good articulation of this characteristic of normal distributions. However, it does not follow that the group of students generally falls within 68% of the total population and follows the  distribution of the general population. 

To illustrate, have a look at the graph below. The 68-95-99.7 cut off points are highlighted for the student group (blue) and the general population (red). Because the mean IQ for the student group (`r round(mean(iq), 1)`) is so much higher than the mean of the large population (100), and the IQs for both populations are symmetrical (approximating normal for the student group), a lot of the students' IQs fall outside the central 68% range of the large population's IQs. In fact, roughly half of the data fall higher than this range because the mean and median of the (approx. normal) student group are one SD (15 points) larger than the mean and median of the large population.     

```{r}
library(ggdist)

tibble(group = "student", iq = iq) |> 
  bind_rows(
    tibble(group = "population", iq = rnorm(1e6, 100, 15))
  ) |> 
  ggplot(aes(x = iq, y = group, fill = group)) + 
  stat_slab(
    aes(fill_ramp = after_stat(cut_cdf_qi(cdf, .width = c(0.68, 0.95, 0.997))))
  ) + 
  theme_minimal() + 
  theme(legend.position = "none")
```

We were looking for an answer that stated that the mean IQ for the student group is higher than that of the large population. The SD for both groups is roughly equal (`r round(sd(iq), 1)` and 15), so they are roughly equally spread out around their means. 

:::

## Question 3

*Points: 5*

::: {.callout-tip title="HG comments"}
Correct. 5/5.
:::

A polling firm is interested in determining how different characteristics of an individual affect vote choice. They record the following characteristics for each survey respondent - race, age, household income, and party affiliation. What "type" of variable does each characteristic most likely represent?

### Answer 
Race and party affiliation represent categorical and nominal variables as they are not quantifiable entities and cant be order. While household income and age represent continuous variables as they can assume an infinite number of real values. 





## Question 4

*Points: 5*

:::{.callout-tip title="HG comments"}
Nice plots. 4/5.
:::

Using the `states` dataset (the unit of analysis is each U.S. state), generate a box plot and density curve for each of the `vep12_turnout` and `cig_tax12` variables. Describe the distribution of each variable in detail.

::: callout-note
The `states` data set can be found in `poliscidata::states`. Take a look at `?states` to see what these variables measure.
:::

```{r}


#  Box Plot and Density Curve of Cigarette Tax per pack in 2012

boxplot(poliscidata::states$cig_tax12,
  ylab = "Cigarette Tax")

ggplot(poliscidata::states, aes(x=cig_tax12)) + 
  geom_density()
```
The variable has a distribution skewed to the left with most of the data centered around 1 with an outlier 

:::{.callout-tip title="HG comments"}
I always mix these up! It's right skewed. 
:::

```{r}
# Box Plot and Density Curve of Voter Turnout of Eligible Population in 2012

boxplot(poliscidata::states$vep12_turnout,
  ylab = "Voter Turnout")

ggplot(poliscidata::states, aes(x=vep12_turnout)) + 
  geom_density() 
``` 
The distribution of the vep12_turnout variable is standard with most of the data  falling within 1 and 2 standard deviations from the mean. 
  
?states

