---
title: "Problem Set 5"
subtitle: "Due date: 23 October"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 25**

```{R}

library(tidyverse)
library(ggplot2)
library(scales)

```

## Question 1

*Total points: 6*

::: {.callout-tip}
4/6
:::

Use the data in the table below to answer the following questions.

```{r}
#| echo: false

tibble::tibble(
  `Age group` = c("18-24", "25-30", "31 and up", "TOTAL"),
  `Non-voters` = c(70, 40, 220, 330),
  Voters = c(50, 50, 570, 670),
  Total = `Non-voters` + Voters
) |> 
  knitr::kable(caption = "Voting by Age in 2000")
```

### Part A

*Points: 2*

::: {.callout-tip}
2/2
:::

What is the probability of being 25-30 or a non-voter?

```{r}

age_25_30 <- 40
non_voters <- 330
total_population <- 1000

probability <- (age_25_30 + non_voters) / total_population

probability


```

The probability of being a non-voter and being aged is 37%

### Part B

*Points: 4*

::: {.callout-tip}
1/4
:::

Assuming a normal distribution, report the 95% confidence intervals for the percentage of 18-to-24-year-olds who did not vote, and then the percentage of 25-to-30-year-olds who did not vote.

```{r}

confidence_interval <- function (p, n) {
  se <- sqrt(p*(1-p)/n)
  lower_range <- p - 1.96*se
  upper_range <- p + 1.96*se
  return(c(lower_range,upper_range))
}

p_18_to_24 <- 70/120
CI_18_to_24 <- confidence_interval(p_18_to_24, 120)

# Careful here! This is p_18_to_24.
p_25_to_30 <- 70/120 
CI_25_to_30 <- confidence_interval(p_25_to_30, 90)


list(CI_18_to_24 = CI_18_to_24, CI_25_to_30 = CI_25_to_30)
```

```         
CI Non-Voters 18 to 24
[1] 0.4951233 0.6715434

CI Non-Voters 25 to30
[1] 0.4814771 0.6851896
```

::: {.callout-tip} 
The probability that an individual is 18-to-24 and a non-voter is not conditional. Therefore, you look at the proportion of the whole population who meet those criteria.

$$
\hat{p} = Pr(18\ to\ 24 \& NV) = \frac{70}{1000} = 0.07 
$$

And: 

$$
n = 1000
$$

Therefore: 

$$
SE(\hat{p}) = \sqrt{\frac{0.07 * 0.93}{1000}} = 0.008068457
$$

The 95% confidence interval can be found:

$$
\hat{p} \pm 1.96*SE(\hat{p})
$$

Therefore, the 95% confidence interval for the percentage of 18-to-24-year-olds who did not vote is `r percent(0.07 - 1.96 * 0.008068457, accuracy = 0.01)` to `r percent(0.07 + 1.96 * 0.008068457, accuracy = 0.01)`.

Using the same method, the 95% confidence intervals for the percentage of 25-to-30-year-olds who did not vote is `r percent(0.04 - 1.96*0.006196773, accuracy = 0.01)` to `r percent(0.04 + 1.96*0.006196773, accuracy = 0.01)`.
:::

## Question 2

*Total points:* *7*

::: {.callout-tip}
7/7
:::

Assume that the standard deviation for the population distribution of a state in which you want to conduct a poll is 200.

### Part A

*Points: 3*

::: {.callout-tip}
3/3
:::

Calculate the spread of the sampling distribution for each of the following sample sizes: 1, 4, 25, 100, 250, 1000, 5,000, and 10,000.

```{R}

population_stddev <- 200

sample_sizes <- c(1, 4, 25, 100, 250, 1000, 5000, 10000)

standard_errors <- population_stddev / sqrt(sample_sizes)

results <- data.frame(Sample_Size = sample_sizes, Standard_Error = standard_errors)
print(results)





```

### Part B

*Points: 1*

::: {.callout-tip}
1/1
:::

Describe specifically how the variability of the sampling distribution changes as the sample size varies. Considering the expense of running a poll, which sample size do you think is most optimal if conducting the poll?

Answer:

A smaller sample size would mean the mean is more likely to deviate from the population mean and the confidential interval would be wider. A larger sample size would mean the sample mean is closer to the population mean and the confidence interval would be narrower and more precise. Given the large costs and the time it takes to conduct a large sample poll, the 1000 number as it would balance the cost and accuracy the most effectively.

### Part C

*Points: 3*

::: {.callout-tip}
3/3
:::

Display your results graphically (using R) with the sample size on the x-axis and the standard error (of the sampling distribution) on the y-axis.

```{r}

library(ggplot2)

sample_sizes <- c(1, 4, 25, 100, 250, 1000, 5000, 10000)
standard_errors <- 200 / sqrt(sample_sizes)  # Calculate standard errors

data <- data.frame(Sample_Size = sample_sizes, Standard_Error = standard_errors)

ggplot(data, aes(x = Sample_Size, y = Standard_Error)) +
  geom_point() +
  labs(title = "Sampling Distribution vs. Sample Size",
       x = "Sample Size",
       y = "Standard Error")




```

## Question 3

*Points: 4*

::: {.callout-tip}
4/4
:::

Suppose you conduct a survey (to generate a sample mean of interest) and find that it has a margin of error of 4.5 with a sample size of 900 using a 95% confidence interval. What would the margin of error be for a 90% confidence interval?

```{r}

E_95 <- 4.5
Z_90 <- qnorm(0.95) 
Z_95 <- qnorm(0.975) 
E_90 <- E_95 * (Z_90 / Z_95)
E_90




```

Answer:

Margin of error for a 90% confidence interval is 3.78

## Question 4

*Points: 4*

::: {.callout-tip}
4/4
:::

Assume that, in State A, the mean income in the population is \$20,000 with a standard deviation of \$2,000. If you took an SRS of 900 individuals from that population, what is the probability that you would get a sample mean income of \$20,200 or greater? What would be the probability if the sample size was only 25?

::: callout-note
Assume a normal distribution for both questions.
:::

```{r}

population_mean <- 20000
population_stddev <- 2000

n <- 900

se <- population_stddev / sqrt(n)

sample_mean <- 20200
z_score_900 <- (sample_mean - population_mean) / se

probability_900 <- 1 - pnorm(z_score_900)

cat("Probability (n = 900):", probability_900, "\n")

n <- 25

se <- population_stddev / sqrt(n)

sample_mean <- 20200
z_score_25 <- (sample_mean - population_mean) / se

probability_25 <- 1 - pnorm(z_score_25)

cat("Probability (n = 25):", probability_25, "\n")




```

A sample of size of 900 the probability of getting a a sample mean income of \$20,200 or greater would be .13%. With a sample size of 25 the probability would be 31%.

## Question 5

*Points: 4*

:::{.callout-tip}
3/4
:::

Assume that a coin is fair. If I flip a coin 500 times, what is a 95% confidence interval for the range of the count of heads that I will get? What if I flip the coin 5,000 times? What about 50,000 times?

Answer for 500 times:

```{r}
n_500 <- 500
prob500 <- 0.5

z_500 <- qnorm(0.975)


CI_95_500 <- c(prob500 -z_500 * sqrt(prob500 *(1 - prob500)/ n_500),
               prob500 + z_500 * sqrt(prob500 * (1 - prob500)/ n_500))


CI_95_500
               







```

Answer for 5000 times:

```{r}

n_5000 <- 5000
prob5000 <- 0.5

z_5000 <- qnorm(0.975)


CI_95_5000 <- c(prob5000 -z_5000 * sqrt(prob5000 *(1 - prob5000)/ n_5000),
               prob5000 + z_5000 * sqrt(prob5000 * (1 - prob5000)/ n_5000))


CI_95_5000


```

Answer for 50000 times:

```{r}

n_50000 <- 50000
prob50000 <- 0.5

z_50000 <- qnorm(0.975)


CI_95_50000 <- c(prob50000 -z_50000 * sqrt(prob50000 *(1 - prob50000)/ n_50000),
               prob50000 + z_50000 * sqrt(prob50000 * (1 - prob50000)/ n_50000))


CI_95_50000





```

::: {.callout-tip}
Almost! The question asks for the confidence interval for the range of the *count* of heads. So, your center point should the number of heads you flipped in these trials. If you have a 50% chance of getting heads when you flip a fair coin, your expected number of heads should be 250, 2,500, and 25,000 heads from each trial. You calculate your confidence around this result. 

```{r}
p_head <- 0.5

tibble(
  flips = c(500, 5000, 50000),
  expected_n_heads = flips*p_head,
  sd = sqrt(flips*p_head*(1-p_head)),
  lower_ci = expected_n_heads - qnorm(0.975)*sd,
  upper_ci = expected_n_heads + qnorm(0.975)*sd
) |> 
  knitr::kable()
```
:::

