---
title: "Problem Set 9"
subtitle: "Due date: 4 December"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 40**

```{r}
library(tidyverse)
```

## Question 1

*Points: 10*

:::{.callout-tip}
10/10
:::

Table 1 below reports the results from two regression models. In Model 1, in Table 1, $Y$ is regressed on $X_1$ and, in Model 2, $Y$ is regressed on both $X_1$ and $X_2$. Why might $X_1$ be statistically significant at conventional levels in Model 1 but statistically insignificant in Model 2? Be as specific as possible.

**Answer:**

In Model 1, $Y$ is regressed on $X_1$​ only, while in Model 2, $Y$ is regressed on both $X_1$​ and $X_2$​. It is possible that $X_1$​ is statistically significant in Model 1 but statistically insignificant in Model 2 due to the presence of multicollinearity between X1​ and X2​. When $X_1$​ is regressed on $Y$ alone in Model 1, it may have a significant effect on $Y$ because it is not being influenced by the presence of $X_2$​. However, when $X_1$​ is included in Model 2 along with $X_2$​, the effect of $X_1$​ on $Y$ may be reduced or eliminated due to the presence of multicollinearity between $X_1$​ and $X_2$.​

## Question 2

*Points: 10*

:::{.callout-tip}
6/10
:::

Using the `censusAggregate` dataset (posted on ELMs) --- which is survey data aggregated to the state level (1972-2000) --- estimate a regression with `vote` as the dependent variable and the following independent variables: `nonSouth`, `edr`, and `pcthsg`. Report the results in a professionally formatted table and interpret the regression results.

Also, create a figure to display the predicted values (with confidence intervals) for the effect of `pcthsg` on the turnout rate. Lastly, is it meaningful to interpret the constant term on its own? Why, or why not?

::: callout-note
`vote` is the turnout rate in a state in a given year (i.e., the number of people who voted divided by the number eligible to vote).

`nonSouth` is a dummy variable equal to `0` for Southern states and a `1` for non-Southern states.

`pcthsg` is the percentage of the population in a state that graduated high school.

`edr` is a dummy variable equal to `1` for states that used election-day registration and a `0` for states without election-day registration.
:::

```{r}

library(poliscidata)
library(stargazer)
library(ggplot2)
library(broom)

censusAggregate <- read.csv("censusAggregate.csv")

model <- lm(vote ~ nonSouth + edr + pcthsg, data = censusAggregate)

stargazer(model, type = "text", title = "Regression Results", 
          dep.var.labels = "Vote", 
          covariate.labels = c("Non-South", "EDR", "Percentage Housing"))

censusAggregate$pred <- predict(model, newdata = censusAggregate, interval = "confidence")
```

:::{.callout-tip}
You need to make sure your problem set renders without error. You needed to remove that code above to prevent it from aborting the render. 
:::

```{r}

library(marginaleffects)

plot_predictions(model, condition = "pcthsg")


```

The coefficient for 'Non-South' is 5.546 with a standard error of 0.797, and it is statistically significant at the 1% level (p\<0.01). The coefficient for 'EDR' is 5.792 with a standard error of 1.056, and it is also statistically significant at the 1% level (p\<0.01). The coefficient for 'Percentage Housing' is 0.101 with a standard error of 0.036, and it is statistically significant at the 1% level (p\<0.01).

:::{.callout-tip}
`pcthsg` is not "percentage housing". From above, `pcthsg` is the percentage of the population in a state that graduated high school.

This points to a common problem with your interpretation. By not translating these variable names into what they represent, you are not describing the phenomenon you are tasked with describing. For example, what is "the coefficient for EDR"?
:::

The constant term represents the expected value of the dependent variable when all the independent variables are zero. It may not make sense to interpret the constant term on its own because it may not be meaningful or realistic to have all independent variables at zero. For example, in this case, having 0% high school graduates for example pcthsg = 0 might not be a realistic scenario. Therefore, the constant term is often used to adjust the regression line, but not interpreted on its own.

## Question 3

*Points: 5*

:::{.callout-tip}
3/5
:::

Using the regression results from the previous question, evaluate the null hypothesis that the effects (i.e., regression coefficients) of `nonSouth` and `pcthsg` are jointly equal to zero. Can you reject the null hypothesis? Be sure to demonstrate how you reached a definitive conclusion.

**Answer:**

The null hypothesis for this test is that both of these coefficients are both equal to zero, which would then mean that these variables have no or minimal effect on the dependent variable Vote. Based on the regression results, both 'Non-South' and 'Percentage Housing' have p-values less than 0.01. This indicates that both of these variables are statistically significant at the 1% level. There is therefore a less than 1% probability that these coefficients are actually zero, given the data we have. Therefore, we can reject the null hypothesis that the effects of 'Non-South' and 'Percentage Housing' are jointly equal to zero. These variables do have a statistically significant effect on 'vote'.

:::{.callout-tip}
To answer this question, you needed to fit the null model (in which the effect of `nonSouth` and `pcthsg` on `vote` is zero). You can then compare the predictive power of that null model to your full model. 
:::

## Question 4

*Points: 15*

::: {.callout-tip}
11/15
:::

Using one of the other datasets available in the `poliscidata` package pick one dependent variable and two or more independent variables. Run a regression of the dependent variable on the independent variables. In your answer, describe why you picked the variables you did, produce a professionally formatted results table, and describe your statistical and substantive findings.

```{r}
data(world)

model <- lm(debt ~ fertility + free_business, data = world)

stargazer(model, title="Regression Results", align=TRUE, type = "text")




```

:::{.callout-tip}
A professionally formatted table includes meaningful variable labels. 
:::

The variables we will for this regression analysis use **`debt`** (public debt as a percentage of GDP) as the dependent variable and **`fertility`** (total fertility rate: number children born per woman) and **`free_busines`** is a business freedom ranking from the heritage foundation. I choose the variables to look at how the number children in a given country and the the freedom to conduct business could correlate with the the amount of public debt a country has accumulated. The coefficient for 'fertility' is -4.738 with a standard error of 2.871.

:::{.callout-tip}
You haven't told me your hypothesis, or what relationship you are testing. You have just told me what your regression is.  
:::


This suggests that for each unit increase in fertility, the debt decreases by 4.738 units, holding all other variables constant. The coefficient for 'free_business' is 0.342 with a standard error of 0.211. This implies that for each unit increase in 'free_business', the debt increases by 0.342 units, with all other variables constant. 

:::{.callout-tip}
Again, I don't know what these variables measure and so I do not understand your interpretation of the regression results. 
:::

The constant term is 36.159 with a standard error of 18.703, which is statistically significant at the 10% level (p\<0.1). This means that when all independent variables are zero, the expected value of 'debt' is 36.159. The R-squared value is 0.076, indicating that approximately 7.6% of the variation in a countries debt can be explained by our IV variables of fertility and business freedom. The F-statistic is 4.660 and is statistically significant at the 5% level (p\<0.05). This suggests that there is a statistically significant relationship between the predictors of fertility and the free business metric and the response variable i.e debt.
